---
title: "Regmods-016 Course Project"
author: "Duncan Turnbull"
date: "17 December 2014"
output: html_document
---


# Executive Summary
This project seeks to answer two key questions around fuel consumption based on car data from the 1974 Motor Trend US magazine: 

+ "Is an automatic or manual transmission better for MPG”. 
+ "Quantify the MPG difference between automatic and manual transmissions"
The data provided had 32 samples with 11 variables. Correlations amongst the variables suggests that weight and the number of cylinders were also variables of interest

Once consideration for weight and engine cylinders are included it does not appear that a significant differnce between manual and automatic transmission can be seen. The primary factors are the car weight and the engine size.

## Look at the data
Looking particularly at how automatic and manual transmissions vary


##1. Identify variables of interest
Potential factors in the data - am (Transmission 0 = auto, 1 = manual)

Look at multi variate model, 
rule out anything check VIF
qsec (1/4 mile time) seems unlikely, maybe drat (rear axle ratio) too  
Other variables probably affect fuel consumption similarly whether manual or auto i.e. weight, cylinders, displacement, horsepower, V/S number of carburetors
The number of forward gears will probably influence the choice

##2 .Final models 
```{r}
fit <- lm(mpg ~ factor(am) +wt ,mtcars)
```

Exploratory Data analyses
Fit multiple models and detail strategy for model selection
Did the student answer the questions of interest or detail why the question(s) is (are) not answerable?
Did the student do a residual plot and some diagnostics?
Did the student quantify the uncertainty in their conclusions and/or perform an inference correctly?
Was the report brief (about 2 pages long) for the main body of the report and no longer than 5 with supporting appendix of figures?
Did the report include an executive summary?
Was the report done in Rmd (knitr)?

#Appendix

## Density / distribution of interesting variable
```{r }
par(mfrow=c(2,2)) ; plot(density(mtcars$mpg));plot(density(mtcars$wt));
plot(density(mtcars$cyl));plot(density(mtcars$am))
```

##MPG by transmission
From the boxplot below it appears as if there are clear distinctions between the MPG for different transmission types. However there could be some confounding variables.

```{r, echo=FALSE}
par(mfrow=c(1,1))
trans <- c("Auto","Manual")[mtcars$am+1]
boxplot( mpg~ trans ,data=mtcars ,main ="MPG by Transmission", ylab="MPG", xlab="Transmission", col = terrain.colors(3))
legend("topleft", inset=.05, title="Transmission", c("Auto","Manual"), fill=terrain.colors(3), horiz=TRUE)
```

##Correlations
The correlation table shows strong correlation to weight (wt), number of cylinders (cyl), size of engines (disp), gross horsepower (hp), however hp, disp and cyl are all strongly correlated with each other suggesting only one factor is required. For our purposes we choose cyl, the number of cylinders.

```{r, echo=FALSE}
cor(mtcars)[c(1:4,6),]
```


##Residuals
The residual plots appear relatively normally distributed and show any observations with undue influence.

```{r, echo=FALSE}
fit <- lm(mpg ~ factor(am) +wt + cyl ,mtcars)
par(mfrow=c(2,2))
plot(fit, col=3 +mtcars$am)
```